---
title: "File Reformatting Example"
author: "James Eddy"
date: "January 26, 2015"
output: word_document
---

First, load the required libraries.
```{r load_libaries, message=FALSE}
library(synapseClient)
library(xlsx)
library(dplyr)
library(knitr)
```

Login to Synapse using credentials saved in .synapseConfig file.
```{r syn_login}
synapseLogin()
```

Using functions in `synapseClient` and `xlsx`, load data from the Synapse
object into R.
```{r get_data, cache=TRUE}
# Define paths for required Synapse objects
unformatted_gwas_covars_address <- "syn3025476" # covariates Excel file

# Download files from Synapse
covars_file <- synGet(unformatted_gwas_covars_address)
covars_file_path <- getFileLocation(covars_file)

# Load the files into R
unformatted_gwas_covars <- read.xlsx2(covars_file_path, 2, 
                                      stringsAsFactors = FALSE)
```

AMP-AD partner teams were provided with a template to follow for providing and
formatting clinical variables. I've used the column headers from this template
to display an example, but I'll also use them to rearrange the Mayo data below.

```{r build_template}
# Define column headers from template
headers <- c("participant_id", "age_at_onset", "age_at_last_assessment", 
             "age_at_death", "post_mortem_interval", "sex", "education",
             "apoe_genotype", "race_ethnicity", "braak_stage", "mmse_at_onset",
             "mmse_at_last_assessment", "cerad")

template <- data_frame("sample1", 75, 86, 86, 10, "M", 12, "E3E4", 
                  "European American", "V", 24, 18, 30)
names(template) <- headers
template[2, ] <- c("sample100", 82, 88, 89, 4, "F", 17, "E2E2", 
                  "African American", "VI", 23, 17, 17)


```

Here's what the template looks like:
`r kable(template, format = "markdown", align = "l")`

For comparison, here is what the Mayo table looks like, before reformatting:
`r kable(head(unformatted_gwas_covars), format = "markdown", align = "l")`

Now for the fun part! I'll rearrange and reformat the data to fit the template,
placing NA entries for any information not included in the Mayo table.

```{r reformat_data, eval=FALSE}
# Simple function to recode APOE genotype value
recode_apoe_status <- Vectorize(function(apoe_status) {
    if (apoe_status == 1) {
        "E4"
    } else if (apoe_status == 0) {
        "none"
    } else {
        NA
    }
})

# Do the reformatting with dplyr (and pipes!)
gwas_covars <- unformatted_gwas_covars %>% 
    # pull out relevant variables from original data frame
    select(participant_id = IID, age_at_diagnosis = AgeOver60, 
           sex = Sex, apoe_genotype = APOE4_Dose) %>%
    
    # rename and modify variables to match template
    mutate(age_at_last_assessment = as.numeric(age_at_diagnosis) + 60,
           sex = ifelse(sex == 1, "M", "F"),
           apoe_genotype = recode_apoe_status(apoe_genotype)) %>%
    
    # add empty columns for additional template variables
    mutate(age_at_onset = NA,
           age_at_death = NA,
           post_mortem_interval = NA,
           education = NA,
           race_ethnicity = NA,
           braak_stage = NA,
           mmse_at_onset = NA,
           mmse_at_last_assessment = NA,
           cerad = NA) %>%
    
    # reorder variables to match template column order
    select(one_of(headers))
```

The following code is used to upload the new, reformatted data back to our
Synapse site.

```{r put_data, eval=FALSE}
# Write new data frame to text file
file_path <- file.path(tempdir(), "mayo_gwas_clinical_vars.txt")
write.table(gwas_covars, file_path)

formatted_covars_file <- File(path = file_path, 
                              parentId = covars_file$properties$parentId)
formatted_covars_file <- synStore(formatted_covars_file)
```

Importantly, I can use my custom `build_relationship` function to define the
provenance for this activity, linking input to code to output. Note: I'll need
to check on Synapse for the object ID of the file I just uploaded.

```{r set_provenance, eval=FALSE}
source("synapseHelpers/R/set_provenance.R")

activity_name <- "clinical data formatting"
input_files <- c("syn3025476")

code_address <- paste0("https://github.com/jaeddy/ampSynapseProjects/",
"raw/master/synapseUploads/R/format_gwas_covars.R")
code_files <- list(list(name = "format_gwas_covars.R",
                        url = code_address))
output_files <- c("syn3026432")
description <- paste("To execute run:", 
                     "Rscript format_gwas_covars.R")

build_relationship(activity_name, input_files, code_files, 
                   output_files, description)
```